---
title: "Shared Ethical Dilemmas For Surgery and Critical Care"
author: "Callum Taylor"
date: "21/June/2021"
output:
  revealjs::revealjs_presentation:
    theme: solarized
    transition: none
    self_contained: false
    reveal_plugins: ["notes"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  include = FALSE
)
source("C:/Users/desktop-coll/Dropbox/Medicine/Academic.9.ICM/surgical_critical_ethics/presentation/diagrams/face_graph_function.R")
```

# Introduction

<div class="notes">



</div>

# Theory

<div class="notes">
</div>

# # The Ethical Systems

Life, Death - Outcomes - "Consequential"

Good, Bad - Values - "Non-Consequential"

<div class="notes">


</div>

# # Principlism - Shared Language

- *Autonomy* - Respecting the person as a person with their own rights
- *Justice* - Treat people fairly
- *Beneficence* - Do good
- *Non-Maleficence* - Dont do bad

<div class="notes">



</div>



# # A Shared Morality?
# # Alternatives
# # Law
# Practice
# # What Are The Chances?


# ## Outcome - Live Vs Die

```{r survival_outcomes}
face_graph_function(
  patient_numbers = 35,
  plot_name = "last_year.svg"
)
face_graph_function(
  patient_numbers = 25,
  plot_name = "hospital_death.svg"
)
face_graph_function(
  patient_numbers = 33,
  plot_name = "survive_year.svg"
)
```

# ### Hospital Inpatient <br> Last Year Of Life (1 in 3){data-background="diagrams/last_year.svg"}

# ### Admitted To ICU <br> Die In Hospital (1 in 5){data-background="diagrams/hospital_death.svg"}

# ### Survive ICU And Hospital <br> Die Within The Year (1 in 5){data-background="diagrams/survive_year.svg"}

<div class="notes">

</div>

# ###  {data-background-iframe="https://callumgwtaylor.shinyapps.io/survival/"}

*[link](https://callumgwtaylor.shinyapps.io/survival/)*


# ### Outcome - Good Life vs Bad Life - The Quality of Life

```{r qol_outcomes}
face_graph_function(
  patient_numbers = 25,
  plot_name = "hospital_readmissions.svg"
)
face_graph_function(
  patient_numbers = 25,
  plot_name = "adls.svg"
)
face_graph_function(
  patient_numbers = 33,
  plot_name = "lost_work.svg"
)
face_graph_function(
  patient_numbers = 50,
  plot_name = "mobility.svg"
)
face_graph_function(
  patient_numbers = 44,
  plot_name = "depression.svg"
)
face_graph_function(
  patient_numbers = 70,
  plot_name = "pain.svg"
)
```

# ### Readmitted To Hospital Within 3 Months (1 in 4) {data-background="diagrams/hospital_readmissions.svg"}
# ### Require Assistance With ADLs at 6 months (1 in 4) {data-background="diagrams/adls.svg"}
# ### Lost Work at 12 Months (1 in 3) {data-background="diagrams/lost_work.svg"}
# ### Need Help With Mobility (1 in 2) {data-background="diagrams/mobility.svg"}
# ### Experiencing Moderate/Extreme Depression at 12 Months (44%) {data-background="diagrams/depression.svg"}
# ### Experiencing Moderate/Extreme Pain at 12 Months (70%) {data-background="diagrams/pain.svg"}

```{r qol_outcomes_two}
face_graph_function(
  patient_numbers = 50,
  plot_name = "lost_work_families.svg"
)
face_graph_function(
  patient_numbers = 25,
  plot_name = "money.svg"
)
face_graph_function(
  patient_numbers = 33,
  plot_name = "ptsd.svg"
)
```

# ## Outcome - Good Life vs Bad Life - The Quality of Life

# ### Families Providing Care  <br> Lose Work (1 in 2) {data-background="diagrams/lost_work_families.svg"}
# ### Families Providing Care <br> Spend Savings/Borrow Money/Sell House (1 in 4) {data-background="diagrams/money.svg"}
# ### Families w/ PTSD after admission (1 in 3) {data-background="diagrams/ptsd.svg"}

# Outcome - Good Life vs Bad Life - The Quality of Life

- HR-QoL Marked Reduction at 12 Months
- Disability Paradox/Hedonic Ladder/Adjustive Preferences

<div class="notes">

Do These Things Matter?

</div>

```{r qol_outcomes_three}
face_graph_function(
  patient_numbers = 70,
  plot_name = "pain.svg"
)
face_graph_function(
  patient_numbers = 50,
  plot_name = "discomfort.svg"
)
face_graph_function(
  patient_numbers = 50,
  plot_name = "grief.svg"
)
```

# Outcome - Good Death vs Bad Death

# ### Pain In ICU (70%) {data-background="diagrams/pain.svg"}
# ### Discomfort In ICU (70%)  {data-background="diagrams/discomfort.svg"}

- ETT
- Hallucinations
- Medical Activity
- Noise/No Rest
- Pain
- Thirst
- Unable to talk
- Shortness of breath
- Being Afraid

# ### Complicated Grief (50%) {data-background="diagrams/grief.svg"}

# Outcome - A Good Death

- Death can be good if it doesn't rob us of future valuable life,
- If it is associated with comfort and consolation
- If it is consistent with the patients values and preferences
- If it is not distressing for the individuals family and for those caring for them [@Wilkinson2020]


# ## ICU Outcomes
# ## Patient Preferences
# An Approach
# MORAL Balance

- Make sure of facts
- Outcomes of Relevence to Agents involved
- Level out the arguments

- in a Balance Box - Using a principlist system

# MORAL Balance Example

> Tracheostomy insertion for 67 year old man with laryngeal carcinoma and impending airway obstruction

[Moral Balance](http://www.moralbalance.org/framework/analyses/)

# Make Sure of Facts

67 year old presented unwell, with weight loss, difficulty swallowing, confusion, agitation and aggression.

<div class="notes">
Delirium diagnosed.
CRP = 158ng/ml but no evident source of probable infection
Peripheral vascular disease
Known vascular dementia
Heavy smoker, heavy alcohol use, said to be ‘always miserable’
Separated from wife 9 years, wife still in contact, also has son and daughter
Living alone, coping progressively less well over 9 months
CT scan chest revealed 5x5cm laryngeal tumour obstructing airway
Resisted IV fluids, antibiotics
Nasendoscopy revealed exophytic tumour.
</div>

# Make Sure of Facts - Options

1) symptom control
2) palliative tracheostomy
3) tracheostomy then biopsy, staging CT, consider for radical laryngectomy or radiotherapy

<div class="notes">
Potential for distressing acute airway obstruction
Cure possible but unlikely, would lose voice
Prospects for rehabilitation poor, unlikely to be able to return home
</div>

# Outcomes of Relevance for Agents 

# Outcomes of Relevance for Agents - Patient

- Lacking mental capacity to decide, paranoid
- Did not want to stay in hospital, wanted to go home, resisting medical and nursing care
- Likely to pull at tracheostomy
- Unhappy with life
- Would not have wanted to lose voice or independence

# Outcomes of Relevance for Agents - Patient's Family

- Avoid distress
- Preserve life but not at all costs
- Respect patient’s views
- Consensus with all family members
- Transfer to hospice, to remain in hospital

# Outcomes of Relevance for Agents - Others

- Staff dealing with difficult behaviours
- Staff anticipating managing acute airway obstruction
- Risk of diagnostic overshadowing (dementia, substance abuse)
- Possibility of resolution of delirium
- Exclude major depression

# Level out in Balancing Box


# Leftovers




# When should we stop?

- What is stopping?
- Why would we stop?

<div class="notes">



</div>

# The Balance

- Good Life
- Bad Life
- Good Death
- Bad Death

<div class="notes">



</div>

# The Tipping Point



<div class="notes">


</div>

# The Law

- Statute
- Case Law

<div class="notes">



</div>

# The Law And Our Dilemma

- Do you have a legal right to life?
- Do you have a legal right to a good life?
- Do you have a legal right to death, or a good death?
    - Is death legally allowed?

<div class="notes">


</div>

# Our Dilemma

<div class="notes">


</div>

# Making the balance

- There are no established answers
- There are no established weights for values

<div class="notes">

</div>

# Autonomy

- Non-capacitous, but does not cooperate with medical treatment and wants to leave hospital
- Avoid dependency, need to move to care home
- Avoid loss of voice, need for PEG feed
- Care for his dog

# Burden (Non Maleficence)

- Medical care resisted by patient, likely to pull at tracheostomy
- Loss of voice, need for PEG feed
- Radical treatment unpleasant and unlikely to succeed
- Delirium may resolve, but dementia will progress and depression unlikely to respond to treatment
- Unlikely to rehabilitate well in view of comorbidities
- Prolonging suffering or distress
- Likely to need care home

# Benefit (Beneficence)

- Preserve life
- Avoid distress if acute airway obstruction
- Allow treatment to resolve delirium
- Allow assessment and treatment of mental state
- Allow work-up for potential radical treatment

# Justice 

- Diagnostic overshadowing, parity of esteem (delivery of physical healthcare to people with MH problems)
- Outcome unlikely to be good in view of comorbidities, resources spent aggressive treatment likely to be better used elsewhere

# Document Decision

> The patient lacks mental capacity to decide, so treatment would be on a best interests basis.
> His expressed, but non-capacitous, wish is against medical intervention.
> Tracheostomy was considered to avoid impending acute airway obstruction, to allow acute medical
> treatment, assessment and possible resolution of mental illness, and time for work-up and consideration
> of definitive cancer treatment.
> Family were keen to preserve his life, but not at any cost
> Family’s primary concern was to avoid distress and respect patient’s previously discussed wishes.

# Document Decision

> Outcome
> Following discussion of options, the pros and cons of each, and the beliefs and priorities that the patient would have brought to the decision, it was agreed not to proceed with tracheostomy insertion, and instead to treat any distress with palliative drugs, including ‘rescue midazolam’ in the case of acute airway obstruction.
> Transferred to a hospice. He became gradually weaker and although his stridor worsened, he died peacefully on day 19, without acute airway obstruction.
