---
title: "Shared Ethical Dilemmas For Surgery and Critical Care"
author: "Callum Taylor"
date: "21/June/2021"
output:
  revealjs::revealjs_presentation:
    theme: solarized
    transition: none
    self_contained: false
    reveal_plugins: ["notes"]
---

<div class="notes">

# Shared Ethical Dilemmas for Surgery and Critical Care

Hi everyone, thank you for sticking around to listen to me, I'm aware this is the last talk of the day.
My names Callum, I'm an intensive care trainee here in Glasgow and I'm currently completing my masters in Medical Ethics and Law

</div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  include = FALSE
)
source("C:/Users/desktop-coll/Dropbox/Medicine/Academic.9.ICM/surgical_critical_ethics/presentation/diagrams/face_graph_function.R")
```

```{r survival_outcomes}
face_graph_function(
  patient_numbers = 35,
  plot_name = "last_year.svg"
)
face_graph_function(
  patient_numbers = 20,
  plot_name = "hospital_death.svg"
)
face_graph_function(
  patient_numbers = 20,
  plot_name = "survive_year.svg"
)
```

```{r qol_outcomes}
face_graph_function(
  patient_numbers = 25,
  plot_name = "hospital_readmissions.svg"
)
face_graph_function(
  patient_numbers = 25,
  plot_name = "adls.svg"
)
face_graph_function(
  patient_numbers = 33,
  plot_name = "lost_work.svg"
)
face_graph_function(
  patient_numbers = 50,
  plot_name = "mobility.svg"
)
face_graph_function(
  patient_numbers = 44,
  plot_name = "depression.svg"
)
face_graph_function(
  patient_numbers = 70,
  plot_name = "pain.svg"
)
```

```{r qol_outcomes_two}
face_graph_function(
  patient_numbers = 50,
  plot_name = "lost_work_families.svg"
)
face_graph_function(
  patient_numbers = 25,
  plot_name = "money.svg"
)
face_graph_function(
  patient_numbers = 33,
  plot_name = "ptsd.svg"
)
```

```{r qol_outcomes_three}
face_graph_function(
  patient_numbers = 70,
  plot_name = "pain.svg"
)
face_graph_function(
  patient_numbers = 70,
  plot_name = "discomfort.svg"
)
face_graph_function(
  patient_numbers = 50,
  plot_name = "grief.svg"
)
```

# Shared Ethical Dilemmas For Surgery and Critical Care

<div class="notes">

So this is the title I was provided, shared ethical dilemmas for surgery and critical care.

To be open right from the start, the reason something is an ethical dilemma, is that you find yourself in a situation where there are at least two reasonable arguments to do different, mutually exclusive things. You're trying to make a decision about the right way to act, and you've got two people you trust, one telling you to turn left, one telling you to turn right.

Ultimately, there isn't necessarily one correct way to act in these scenarios.

What this talk aims to do is give you a bit more information with regards to a particular dilemma, that of when to stop intervening with what we call "life sustaining treatments"


I think this talk has three main points:

1 - That the principles we were told were pillars of medical ethics, are not as solid as we were taught.
2 - That our patients are a lot closer to the end of their lives then we expect, when they undergo a critical illness, or major surgery. In addition to that, you are a lot more unwell after critical illness and major surgery than we often acknowledge.
3 - And lastly, that when we give our patients accurate information about their illness and our treatments, they make different decisions, they are more likely to refuse certain treatments, and they may have a better quality of life as a result. I think as surgeons, you are best placed to start these conversations with your patients, especially your elective ones, through the relationship and rapport you have developed

</div>

# Theory

> "The dilemma at the base of intensive care is the tension between a right to life and a right to die"

*Hard Questions in Intensive Care. Dunstan. 1985*

<div class="notes">

So that's the gist of the talk, you can stop listening now if you've had enough.

Dunstan in 1985 summed up what we're talking about here, that we have competing duties, to protect someones right to life, and their right to a good death.

</div>

# Ethical Systems 

Outcomes Versus Values

<div class="notes">

So how should we look at these duties from an ethical framework.

The first division in approach is whether we think it's the outcomes that are important (the good life and good death), or the values that drove us in a certain way (the duties)

If you think the outcomes, you're thinking in consequentialist terms.

</div>

# 

<section data-background-video="diagrams/utility.mp4"
          data-background-video-loop=true data-background-video-muted>
<div class="notes">

So the traditional consequentialist approach would be utitilitarism. The right thing to do is the way that produces the most of a certain outcome.
In our situation that may be act in the way that saves the most lives. It could be any consequence, it could also be a combination of consequences.


</div>
</section>

# 

<section data-background-video="diagrams/utility2.mp4"
          data-background-video-loop=true data-background-video-muted>
<div class="notes">

But we're operating in an uncertain system. You don't know if a patient is going to survive, or how long they'll survive, or what their quality of life may be.

So instead we need to talk about the chances of a certain outcome, or expected utility. They way to work what the best action is, is it's the one most likely to give us the outcome we want, like survival from critical illness.

The problems that start to emerge is that for a strict utilitarian, the only thing that matters is the outcome. What you do to achieve it doesnt matter. The ends justify the means.


</div>
</section>

# Non-Consequentialist 

<div class="notes">


Instead you may want to take on different factors, take on a non-consqequentilist or deontological approach. It's the values or duties that you followed that means you acted in the correct way. You wanted to respect a patients human rights, or to treat them with compassion, or to act in a way you think a good surgeon should act.

At this point like me, you might say, this traditional split between only outcomes or only reasons seems daft, clearly they both matter.
</div>

# Principlism

- *Autonomy* - Respecting the person as a person with their own rights
- *Justice* - Treat people fairly
- *Beneficence* - Do good
- *Non-Maleficence* - Dont do bad

<div class="notes">
And this is where the approach of Beauchamp and Childress comes in.

I don't know for everyone here, but I was taught this in uni as the pillars of medical ethics. These four principles: of respecting the person, treating everyone fairly, doing good, and avoiding harm (or autonomy, justice, beneficence and nonmaleficence) were the foundations of good ethical decisions.

What principlism does is say rather than one absolute factor is important, instead theres this plurality of factors that need to be balanced against one another.

Beauchamp and Childress put it forward that these four principles are universal, that they come from our shared common morality. It's these four principles, because we all already know that it's these four principles. It's here that the difficulty emerges.

```{r, echo= FALSE}
DiagrammeR::mermaid("
graph TD;
A[Principlism]
A---B[Autonomy]
A---C[Justice]
A---D[Beneficence]
A---E[Non Maleficence]
", height = 1000)
```
</div>

# Shared Morality? - House of Sand 

> Based on our findings, we can conclude that the principles of biomedical ethics – in particular autonomy – have only a weak grounding in human moral psychology and thus in the common morality 

*Christen 2014*

<div class="notes">
So it's been reasonably criticised, their idea of a common morality. The idea that the two affluent white male north american academics have a perspective that is global, and covers all of our shared morality, seems unlikely. Theres culture and sociology that comes into what certain groups find to be principles of an ethical decision. Walker and Lovat point out that from a sociological perspective, principlism puts autonomy so high as it came about in the late 20th century, when our failings to protect autonomy in nazi experimentation, and the tuskagee trials, was so clear.

If you look at a psychological grounding for the common morality, you don't come up with the four principles.

Christen in 2014 when looking at those in medicine and in business found the two groups had two different moral systems, and neither was the four principles. The parts that overlapped seemed to be: respect, loyalty, and responsibilty.

Now that doesn't mean that these four principles aren't important, but it does suggest that they can't be justified as coming from a common morality like the authos of principlism put forward.

The way I see it. There’s a great big mess of ethical factors that could be relevant to our decision. The four principles approach doesn’t tell us anything fundamental about these factors. What it does though, is provide a nice sorting system for us to group aspects together and compare them against each other. I’d describe the four principles as a filling system, rather than the pillars we get taught.

Principlism is a shared language in medicine. It doesn’t really matter that it’s maybe not grounded in what Beauchamp and Childress say it is, it’s still what we all use. Interestingly, the piece from Christen does suggest that maybe medical practitioners came closer to the four principles than those in business did, as the medics had been taught the four principles in uni, and that had shaped their ways of thinking.

We aren’t going to get away from principlism, and I don’t think we need to. It’s a good labelling system.
</div>

# Human Rights Act

- Article 2 Right To Life
- Article 3 Right to protection from torture (mental or physical), inhuman or degrading treatment or punishment 

<div class="notes">

I won't talk about the law too much, beyond point out that if a dilemma gives us a range of possibly acceptable ways to act, the law provides us with boundaries at either side, at where society currently finds actions to be acceptable.

The classic case in UK law when it comes to stopping life sustaining treatments would be that of Airedale vs Bland. ony Bland was a young man critically injured in the Hillsborough football disaster, and the case was establishing the right to withdraw artificial nutrition and hydration that was keeping him alive in a persistent vegetative state. So the law places a guide here, there isn’t an obligation to pursue the prolongation of life at all costs.

These two articles of the Human Rights Act of 1998 are up on the screen to show that the law faces its own dilemmas. The state has a duty to protect an individuals right to life, but also to protect them from inhuman or degrading treatment. The two of these may come into conflict with aspects of intensive care. In addition Article 5 puts a duty on the state to protect an individuals right to liberty and freedom from detention, something the chemical and physical restraints of surgery and intensive care can interfere with.

</div>

# Practice - Situation

- A Good Life
- A Bad Life
- A Good Death
- A Bad Death

<div class="notes">
So going back to our dilemma, when should we stop, or not start, life sustaining treatments?

We can imagine that whatever we do, one of these four options will be the outcome for the patient:

Our choices, our interventions, can affect the likelihood of which outcome we get. Some of these will be desirable and some won’t be. However the labelling of any outcome is going to depend on the individual affected by it. What makes a life good or not is uncertain. What makes a death good or not is uncertain. How desirable they are is uncertain.

There is at some unknown point for every patient, a tipping point in chance where an intervention they consider worth it, becomes one they don’t want to go ahead with. That may be because it doesn’t give them a good enough chance at their good outcome. Or it may be because it gives too high a chance of a bad outcome. Where the tipping point is for the patient is unlikely to be the same as where we place it..
</div>

# What Are The Chances?

<div class="notes">
In determining where this tipping point is, we need to have our best guess at what is likely to occur. Of course we also need to be honest that sometimes our best guess is pretty rubbish.
</div>

# Hospital Inpatient <br> Last Year Of Life (1 in 3){data-background="diagrams/last_year.svg"}

<div class="notes">
So this is the first point that I'd like to bring. The patients that we meet are more likely to be close to the end of their life than we think.
These statistics are from FICM's Working Group on End of Life care.
</div>

# Admitted To ICU <br> Die In Hospital (1 in 5){data-background="diagrams/hospital_death.svg"}

<div class="notes">
Of those who we consider the potential benefits are worth it to try a spell in ICU:
One in five of them don't survive hospital.
</div>

# Survive ICU And Hospital <br> Die Within The Year (1 in 5){data-background="diagrams/survive_year.svg"}

<div class="notes">
Of those who do survive to get home, one in five of them are dead within the year. And these two points are of the patients who we've already thought are "well enough" for ICU.
</div>

# Readmitted To Hospital Within 3 Months <br> (1 in 4){data-background="diagrams/hospital_readmissions.svg"}

<div class="notes">
Of these patients who survive ICU, and get back home, a quarter are back in hospital within 3 months.
</div>

# [Consolidation](https://callumgwtaylor.shinyapps.io/survival/){data-background-iframe="https://callumgwtaylor.shinyapps.io/survival/"}

<div class="notes">
To consolidate this, we'll guess some chances and take some from scottish cohort data.
If we have our high risk cohort and operate on 100 of them.
Say 95% survive their operation, and 90% of them survive their ICU stay, and 88% of them survive their hospital stay, and 87% of them survive one year, and 73% of them survive the next three years to get to four years post op. Then less than half of the patients we started with are alive at 4 years. It was only 2/3 who made it to a year. Those last few numbers that seemed all to specific, are from a scottish study on survival after surgical stays in critical care.

*************
Note, you had to edit the html here to get the background to work! You had to put the data background inside the h1 tag

</div>


# Require Assistance With ADLs at 6 months <br> (1 in 4){data-background="diagrams/adls.svg"}

<div class="notes">
1 in 4 patients will need assistance with ADLs at the 6 months after discharge mark.
</div>

# Lost Work at 12 Months <br> (1 in 3){data-background="diagrams/lost_work.svg"}

<div class="notes">
1 in 3 will have less work at 12 months
</div>

# Need Help With Mobility <br> (1 in 2){data-background="diagrams/mobility.svg"}

<div class="notes">
1 in 2 will need some form of mobility aid at 6 months
</div>

# Experiencing Moderate/Extreme Depression at 12 Months <br> (44%) {data-background="diagrams/depression.svg"}

<div class="notes">
Over a third, close to half will have symptoms of moderate or extreme depression at 12 months
</div>

# Experiencing Moderate/Extreme Pain at 12 Months <br> (70%) {data-background="diagrams/pain.svg"}

<div class="notes">
The majority will have symptoms of moderate or extreme pain at 12 months
</div>

# Families Providing Care  <br> Lose Work (1 in 2) {data-background="diagrams/lost_work_families.svg"}

<div class="notes">
Half of families will have lost income at the 12 month mark, either due to family members cutting back work to provide care, or related to the patient's lost income
</div>

# Families Providing Care <br> Spend Savings/Borrow Money/Sell House <br> (1 in 4) {data-background="diagrams/money.svg"}

<div class="notes">
Of those families that are providing care. 1 in 4 either have to spend savings, borrow money, or sell their house.
</div>

# Families w/ PTSD after admission <br> (1 in 3) {data-background="diagrams/ptsd.svg"}

<div class="notes">
ICU is not benign for families. 1 in three families will have a relative with PTSD as a result of the ICU admission
</div>

# Summarise Our Situation - Do These Things Matter?

<div class="notes">
So these things are all about survival. There's good evidence that when it comes to health related quality of life measures, that patients have a lower "Quality of life" after surviving critical illness .

But, these are the patients who survived. When you look at whether these patients consider it worth it, they generally say it was. The choice after all is between a lower quality of life or death.

What am I trying to get across here?

Probably that we can over estimate the likelihood that someone will survive intensive care.

Probably that we over estimate the amount that people will be able to do if they survive intensive care.

Probably that we under estimate the harms of intensive care

Probably that most measures of quality of life, under estimate how satisfied someone is with their life.

So Back To The Dilemma. We're looking here at the question of when is enough intervention enough? Which of the outcomes do we think our intervention will result in, a good life or bad life, a good death or bad death. Are the chances of any of these unacceptably high or unacceptably low? We've got a bit more knowledge now to adjust the chances of these.
</div>

# Critical Illness and Death - The Experience, Patient Preferences

<div class="notes">
This last bit is looking at how patients' decisions change with more accurate information.
</div>

#

> patients who did not receive information on prognosis perceived a much longer life expectancy... In addition, these patients were less likely to perform advanced planning and more likely to select aggressive treatment over comfort-oriented care

[https://link.springer.com/article/10.1245/s10434-016-5564-x#Sec3]

<div class="notes">
Patients are going to hope for the best when it comes to their condition and survival. Without accurate information they will assume they have a better chance, or longer, then they actually have.

This assumption that they're in a healthier place will result in them being more likely to pick more aggressive treatments.
</div>

#

> In one study of 590 patients with metastatic cancer, Enzinger et al. reported that the majority (71 %) of patients wished to be informed of their prognosis; despite this, the authors noted that only 18 % actually received an estimate of life expectancy

[https://link.springer.com/article/10.1245/s10434-016-5564-x#Sec3]

<div class="notes">
This is not because patients with life limiting illnesses are attempting to hide away from the truth. The majority want to know. It's us who aren't telling them.
</div>

#

> In a separate prospective longitudinal cohort study of 332 patients with advanced cancer and caregivers, Wright et al. noted that end-of-life discussions did not lead to worry or depression but instead resulted in less aggressive medical care, longer hospice stays, and better quality of life among both patients and caregivers

[https://link.springer.com/article/10.1245/s10434-016-5564-x#Sec3]

<div class="notes">
We often don't tell them the reality due to this fear of removing all hope, of worsening the situation. Now this seems like a reasonable concern, and their are certainly ways that such a conversation could harm a patient. However, when a conversation occurs, it results in a better quality of life for a patient, without worsening depression.
</div>

# 

> Patients with inaccurate prognostic awareness were significantly more likely than those with accurate understanding to prefer ICU care, intubation, and mechanical ventilation than to reject these LSTs.

[https://pubmed.ncbi.nlm.nih.gov/26386187/]

>  Patients express a wide variety of preferences to limit treatments and avoid unacceptable outcomes that have not been discussed with their surgeon.

[https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4949135/]

<div class="notes">
Patients are going for these more aggressive treatments, as mentioned before, as they're starting from a place thinking they are healthier than they are. When patients are told the likelihood of their treatment being successful reducing, unsurprisngly they are less likely to pick it.
</div>

# 

>  The proportion of patients choosing LSTs decreased in line with an increase in the likelihood that the outcome of the treatment would be death or functional and/or cognitive impairment. 

[https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7795676/]

<div class="notes">
This again, is just showing, the more patients know about their treatments, the less likely they are to pick certain options

</div>

#

>  Patients who reported having discussed their wishes for EOL care with a physician were more likely to receive care that was consistent with their preferences, both in the full sample and among patients who were aware they were terminally ill

[https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2834470/]

<div class="notes">
As mentioned above, talking with clinicians about what patients want, meant they were more likely to get treatments that go along with their values
</div>

#

> Patients who recognized their terminal illness were more likely to prefer symptom-directed care (83%). However, some patients who were aware they were terminally ill wished to receive life-extending care (17%, 21 of 121 patients).

[https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2834470/]

<div class="notes">
Of course, I don't want to pretend that every patient wants us to stop. Not even every patient who knows they are dying wants to stop. Some will see benefits to treatment and desire to contine. 17% in that paper above. That raises it's own dilemma.

And understandably, the majority of patients who have multimorbidity, but not necessarily a life limiting illness, will desire life sustaining treatments regardless of the chance of success. Again though, the chance of someone refusing that treatment goes up with the level of information they are provided.
By giving someone full information about the potential outcomes of any intervention, you are more likely going to give them a treatment in keeping with their own values and beliefs
</div>

# Why might an ICU Death be a "Bad Death"

<div class="notes">
So we've talked about the aftermath of intensive care, but why might it be that a death in intensive care is a bad option. Patients don't report it as somewhere they would like to die. And of course we can't ask the patients  who did die, but we can look at the patients experience of ICU by asking those who survived, and making the assumption that those who died had a similar experience
</div>

# Discomfort + Pain In ICU <br> (70%){data-background="diagrams/discomfort.svg"}

- ETT
- Hallucinations
- Medical Activity
- Noise/No Rest
- Pain
- Thirst
- Unable to talk
- Shortness of breath
- Being Afraid

<div class="notes">
The majority of patients who survive ICU describe pain and discomfort in the ICU, from these causes
</div>

# Complicated Grief <br> (50%) {data-background="diagrams/grief.svg"}

<div class="notes">
In addition to that, beyond the patients experience, 50% of families whose relative dies in ICU describe symptoms of complicated grief a year later.
</div>

# What might be a good death?

- Death can be good if it doesn't rob us of future valuable life,
- If it is associated with comfort and consolation
- If it is consistent with the patients values and preferences
- If it is not distressing for the individuals family and for those caring for them [@Wilkinson2020]

<div class="notes">
But what's the alternative? A death's a death, surely they're all as bad as each other?

Wilkinson, a PICU consultant put forward this set of criteria in a paper last year about how we might consider a death a good death, if it doesn't rob us of future life, etc etc. You can invert them to look again at what may be a bad death, and see that that links in pretty well with what an ICU death is.
</div>

# [MORAL Balance](http://moralbalance.org/framework/analyses/)

**M**ake sure of facts

**O**utcomes of **R**elevance for **A**gents

**L**evel out the arguments

in a **Balance** box

<div class="notes">
I thought after all the vagueness we should try and provide something useful as a system. The Moral Balance system has been put forward by intensivists working in the UK, as a way to gather and structure your thoughts when attempting to make an ethical decision in medicine, using the four principles.

Ultimately, what we are doing is formalising what we've been talking about here. In discussion with the patient, their relatives, and your colleagues, you are working out what is important to everyone. You're then seeing how these factors come up against one another, and trying to work out how to either achieve the most important of them, or not cause the greatest harms.

These could be recognising a patients autonomy, even in situations where they may lack capacity, they are still able to express a preference. These can be aiming for beneficence through presevivng life, whilst avoiding non-malefiecence (pain, discomfort, particular interventions a patient had previously refused)
</div>

# Moral Balance Example {data-background-iframe="MORAL-Balance-Analysis-1--Tracheostomy-for-laryngeal-cancer.pdf"}


# Summary


<div class="notes">
So again, ethical dilemmas are dilemmas because there are often multiple valid ways, or ways that appear valid, to approach a situation. There isn't one right answer as to whether to stop life sustaining treatment, and it's very situational dependent. The principles we have been taught can be useful to label the factors we feel are relevant, but they are not the totality, and we shouldn't ignore aspects just because they aren't one of the four pillars.

Our patients are a lot closer to their death then we often acknowledge, and survivng critical illness is difficult, and often comes with long term consequences. The more patients know about their treatment the less likely they are to go through with it, and in some studies thats been associated with a higher quality of life in those groups.

What we can say, is that if you get closer to an individuals own perspective on their life, you are more likely to end up in a scenario in keeping with their values. Patients desire conversations about their prognosis and treatment and if handled well, they help rather than harm. We can only make decisions with patients in keeping with their values, if we know what's important to them, and they know the reality of their treatment.

</div>

# References

- Care At The End Of Life - A guide to best practice, discussion and decision-making in and around critical care, The Faculty of Intensive Care Medicine, 2019
- An exploration of social and economic outcome and associated health-related quality of life after critical illness in general intensive care unit survivors: a 12-month follow-up study, Critical Care, 2013
- How “moral” are the principles of biomedical ethics? – a cross-domain evaluation of the common morality hypothesis, BMC Medical Ethics, 2014
- Discomfort and factual recollection in intensive care unit patients, Critical Care, 2004
- Pain in patients in critical condition and its environment.Revista Médica del Hospital General de México. 2018
- Life and Death Decisions in the Clinical Setting. Walker. 2017
- Intensive care utilization and outcomes after high-risk surgery in Scotland: a population-based cohort study. British Journal of Anaesthesia. 2017